null

### 2017-03-30 Manuscript/Manuscript_COD_Trends.Rmd
more writing

---
title: 
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

@@

###Introduction
The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this figure are differences by race. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.

@@

**The data**
We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.

@@

**Statistical Methods**
The data structure produced two statistical challenges that were addressed by our quantitative approach. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

@@

@@

@@

**Need to add information on the trends in mortality**

###Results

>>>>

---
title: "null"
bibliography: bibliography.bib
output:
  word_document: default
  html_document: default
always_allow_html: yes
---

@@

###Introduction

The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this figure are differences by race. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.

@@

**Description of the mortality data**

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.

@@

**Statistical challenges to be addressed**

The data structure produced two statistical challenges that were addressed by our quantitative approach. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

**Estimation of smoothed mortality rates using a Bayesian autoregressive model**

@@

**Decomposition of the life expectancy gap by cause of death**

@@

**Description of the trend in each cause of death's contribution to the gap and identification of divergent states**

@@

*Need to add information on the trends in mortality* -- where?

###Results
```{r initialize, echo=FALSE, warning=F, message=F}
library(plotly)
library(viridis)
library(scales)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
library(splines)
library(broom)
library(tidyverse)

source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r load_data, echo=F}
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")
BlackWhite_results <- reorder.as.map(BlackWhite_results, "state")

cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
cod_decomp_results$COD_cont_yrs_width <- cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl
cod_decomp_results$COD_cont_yrs_SE <- (cod_decomp_results$COD_cont_yrs_ucl - cod_decomp_results$COD_cont_yrs_lcl)/3.92
```

```{r life-expectancy-trends, echo=F}
LE_trend_plot_males <- ggplot(data = subset(BlackWhite_results, sex == "Male"), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")

LE_trend_plot_females <- ggplot(data = subset(BlackWhite_results, sex == "Female"), aes(y = LE_wbgap_mean, x = year)) + 
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +             
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, fill = Census_Division)) +
  #geom_line(aes(col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)")
```

```{r plot_both, eval=F, echo=F}
#Informative: male and females at the same time
ggplotly(ggplot(data = subset(BlackWhite_results), aes(y = LE_wbgap_mean, x = year)) + #, sex == "Male"
  geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +
  geom_ribbon(aes(ymin = LE_wbgap_lcl, ymax = LE_wbgap_ucl, group = sex, fill = Census_Division), alpha = 0.5) +
  geom_line(aes(group = sex, col = Census_Division)) + #add this for the online version, and then adjust legend and hover display
  facet_wrap(~ state.map.order, ncol = 11, drop = F) +
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_blank(),
        strip.background=element_blank(),
        axis.line=element_blank(),
        axis.ticks=element_blank()) +
  ylab("Life expectancy gap (years)") +
  xlab("Year (1969-2013)"))
```

```{r meta_regression_of_cod_contribution, echo = F}
#run 12 stratified models. bs() created the b-spline basis
models <- cod_decomp_results %>% 
  group_by(COD, sex) %>%
  do(fit = lm(COD_cont_yrs_mean ~ bs(year, degree = 1, df = 5) + state, 
              weights = 1/(COD_cont_yrs_SE)^2, 
              data = .))

#calculated predicted values for all models
dfPred = broom::augment(models, fit)
dfPred$year <- 1969:2013

#merge the predictions with original dataset
cod_decomp_results <- merge(cod_decomp_results, 
                            dfPred %>% select(COD, sex, year, state, .fitted, .resid), 
                            by = c("COD", "sex", "state", "year"))

#calculated the sum of the residuals for each state-sex-cod strata (summed over years)
cod_decomp_results <- cod_decomp_results %>% 
  group_by(state, sex, COD) %>%
  mutate(sum.residual = sum(abs(.resid)))
```

```{r plotting_resid_fx, echo = F}
plot.resid.trends <- function(dataset, COD1, sex1) {
  subset.data <- subset(dataset, COD == COD1 & sex == sex1)
  subset.data$order <- reorder(subset.data$state, subset.data$sum.residual, first)

  a <- ggplot(subset.data,
         aes(x = year, y = COD_cont_yrs_mean)) + 
    #geom_line(aes(col = "Observed")) + 
    geom_ribbon(aes(ymin = COD_cont_yrs_lcl, ymax = COD_cont_yrs_ucl), fill = "#abd9e9") +
    geom_hline(yintercept = 0, col = "black") +
    #scale_fill_manual(values = c("#b2e2e2")) +
    facet_wrap( ~ order) + 
    geom_line(aes(y = .fitted), col = "#d7191c") +
    #scale_color_manual(values = c("#006d2c")) +
    xlab(" ") + ylab("estimated smoothed contribution vs. common trend (years)") + 
    ggtitle(paste0("Estimated smoothed contribution vs. meta-estimated common trend in ", COD1, " in ", sex1)) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

  b <- ggplot(subset.data,
              aes(x = year, y = .resid)) + geom_line(aes(col = "Residual")) + 
    scale_color_manual(values = c("#d7191c")) +    
    facet_wrap( ~ order) +
    xlab(" ") + ylab("Residual") +
    geom_hline(yintercept = 0, col = "black") +
      geom_text(data = subset(subset.data, year == 2010), 
                aes(y = 1, label = round(sum.residual, 1)), size = 2.5) + 
    ggtitle(paste0("Residuals vs. year in ", COD1, " in ", sex1)) +
    theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

    return(list(plot1 = a, plot2 = b))
}
```

```{r create-the-resid-plots, echo = F, message = F}
for(COD_i in levels(cod_decomp_results$COD)) {
  for(sex_i in levels(cod_decomp_results$sex)) {
    assign(paste0(COD_i, ".", sex_i, ".plot"), plot.resid.trends(dataset = cod_decomp_results, 
                                                                         COD1 = COD_i, sex1 = sex_i))
  }
}
```

**Figures 1 and 2** display trends in the life expectancy gap for males and females, respectively. For both sexes, the life expectancy gap has decreased since 1969, although several states exhibited a period of increasing inequality that peaked in the early 1990s, especially among males.

**Figures 2 - 14** display trends in the contribution of each cause of death to the life expectancy gap for males and females, in years. The blue band illustrates the credible interval for the smoothed contribution estimated for each state, and illustrates the heterogeneity across states in their contributions trends. These curves are overlaid with the meta-estimated common trend, shown with a red line, and the plots are organized according to the closeness of the [fit] (*need another word*) of the red curve to the blue band. Thus, increasingly divergent states are shown further down the display. We supplement these main results with smoothed estimates of the age-standardized mortality rates for each cause of death (**eFigures 1-12**).

**Cardiovascular disease**

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for both sexes and races (**eFigures 1 and 2**). However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**). In particular, Pennsylvania, Michigan, Wisconsin, and Washington DC experienced relatively larger increases to their CVD contributions to the gap for males, while some states such as Illinois and Kentucky exhibited relatively larger early contributions in the 1970s.

The situation was reversed for females, with black females experiencing relatively larger mortality decreases over time compared to white females. Thus, the contribution of CVD to the life expectancy gap decreased by more than a year in several states (**Figure 4**). Of note, some states experienced a period of early increasing contributions, most notably, Arizona, Washington, and Oregon, that was not captured by the general pattern. Furthermore, Wisconsin and Washington DC both exhibit increasing contributions over time, as mortality gains among white females outpaced those among black females.

**Cancer**

For all races and sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**). Among males, this inverted U-shape pattern was accentuated among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**). Notable exceptions to this general pattern include Washington DC and Mississippi, states that underwent increasing cancer contributions over time, and New Mexico and Rhode Island, exhibiting decreasing contributions over time (although imprecisely estimated).

For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern is similar between blacks and whites and most states do not exhibit the accentuated period of increase in mortality that was estimated for black males. Thus, the contribution of cancer to the life expectancy gap is generally flat over time (**Figure 6**). Washington DC deviates from this pattern, and exhibited increasing cancer contributions to the gap, while Rhode Island exhibited the reverse. Most notably, some states exhibited no inequality in cancer mortality (**eFigures 3 and 4**), including New York, and Connecticut (and for males, these states recently eliminated the black-white difference in cancer mortality). Furthermore, Massachusetts and (imprecisely) Rhode Island exhibit lower cancer rates among black women and men compared to whites in 2013.

****

**Figure 1: State-level trends in the difference in life expectancy between blacks and whites in males, United States, 1969-2013**

```{r plot_LE_trends_males, echo=F, fig.height=9, fig.width=11}
ggplotly(LE_trend_plot_males)
```

**Figure 2: State-level trends in the difference in life expectancy between blacks and whites in females, United States, 1969-2013**

```{r plot_LE_trends_females, echo=F, fig.height=9, fig.width=11}
ggplotly(LE_trend_plot_females)
```

**Figure 3: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in males**

```{r print-plot, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cardiovascular.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cardiovascular.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
```

**Figure 4: Smoothed state-level trends in the contribution of cardiovascular disease to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-2, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cardiovascular.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cardiovascular.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
```

**Figure 5: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in males**

```{r print-plot-3, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cancers.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cancers.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
```

**Figure 6: Smoothed state-level trends in the contribution of cancer disease to the life expectancy gap vs. the overall pattern in females**

```{r print-plot-4, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cancers.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Cancers.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))
```

```{r print-plot-5, fig.width = 9, fig.height=9, echo = F, eval = F}
ggplotly(Communicable.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Communicable.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(Communicable.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Communicable.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(`Non-communicable.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`Non-communicable.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(`Non-communicable.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`Non-communicable.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(Injuries.Male.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Injuries.Male.plot[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(Injuries.Female.plot[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(Injuries.Female.plot[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(`All other causes.Male.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`All other causes.Male.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))

ggplotly(`All other causes.Female.plot`[[1]]) %>% layout(xaxis = list(title = "Year"))
ggplotly(`All other causes.Female.plot`[[2]]) %>% layout(xaxis = list(title = "Year"))
```

###Discussion

###Appendix
